---
title: "Fast Design of High Order Portfolios"
author: |
  | Rui Zhou and Daniel P. Palomar
  | The Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: vignette
    toc: yes
    toc_depth: 2
  pagedown::html_paged:
csl: ieee.csl
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Design of High-order Portfolios}
  %\VignetteKeyword{High-order portfolio}
  %\VignetteKeyword{skewness}
  %\VignetteKeyword{kurtosis}
  %\VignetteKeyword{portfolio design}
  %\VignetteKeyword{successive convex approximation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knit_hooks$set(pngquant = hook_pngquant)  # brew install pngquant
options(width=100)
# rmarkdown::render("vignettes/DesignOfHighOrderPortfolios.Rmd", "prettydoc::html_pretty")
```

-----------
> This vignette illustrates the design of risk parity portfolios, which are
  widely used by practitioners in the financial industry, with the package
  **highOrderPortfolios**, giving a description of the algorithms and
  comparing the performance against existing packages.


# Quick Start

Let's start by loading the package:
```{r, echo=TRUE}
library(highOrderPortfolios)

?design_MVSK_portfolio  # to get help for the function
```

The simplest use is for the [mean-variance-skewness-kurtosis portfolio](#mvsk-portfolio):

```{r, echo=TRUE}
# load the embedded dataset
data(X50)
str(X50)

# estimate moments
X_moments <- estimate_moments(X50, adjust_magnitude = TRUE)

# decide problem setting
w0 <- rep(1/50, 50)
w0_moments <- eval_portfolio_moments(w0, X_moments)
d <- abs(w0_moments) 
kappa <- 0.3 * sqrt(w0 %*% X_moments$Sgm %*% w0)

# portfolio optimization
sol <- design_MVSKtilting_portfolio(d, X_moments, w_init = w0, w0 = w0, 
                                    w0_moments = w0_moments, kappa = kappa, 
                                    ftol = 1e-10)
```

Let's take a look at the result:
```{r, echo=TRUE}
# plot
require(ggplot2)
require(magrittr)

data.frame(idx = 1:ncol(X50), weights = sol$w) %>%
  ggplot(aes(x = idx, y = weights)) +
  geom_bar(stat = "identity") +
  labs(x = "asset indexes", y = "portfolio weights")
```



# What is a High-Order Portfolio?
## Signal model
Let us denote by $\mathbf{r}_{t}$ the vector of the **returns** of $N$ assets at time $t$ and suppose they follow an i.i.d. distribution
(which is not a totally accurate assumption, but it is widely adopted, nonetheless) with mean $\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$.

The portfolio vector $\mathbf{w}$ denotes the normalized dollar weights allocated to the $N$ assets, and the
**portfolio return** is then $r_{t}^{\sf portf} = \mathbf{w}^{T}\mathbf{r}_{t}$,
with expected return $\mathbf{w}^{T}\boldsymbol{\mu}$ and variance $\mathbf{w}^{T}\boldsymbol{\Sigma}\mathbf{w}$.

## Modern portfolio theory
In 1952, Markowitz proposed in his seminal paper [@Markowitz1952] to find a trade-off between the portfolio expected return and its risk measured by the variance:
$$\begin{array}{ll}
\underset{\mathbf{w}}{\textsf{maximize}} & \mathbf{w}^{T}\boldsymbol{\mu}-\lambda\mathbf{w}^{T}\boldsymbol{\Sigma}\mathbf{w}\\
\textsf{subject to} & \mathbf{w} \ge \mathbf{0}, \quad\mathbf{1}^T\mathbf{w}=1,
\end{array}$$
where $\lambda$ is a parameter that controls how risk-averse the investor is.

Markowitz's portfolio has been heavily critized for over half a century and has never been fully embraced by practitioners, among many reasons because:

- Mean and variance of the portfolio return are good measures when it is Gaussian distributed. But the empirical distribution of the portfolio return is usually asymmetric and heavy-tailed, which can not be well described by only mean and variance.
- The objective $\mathbf{w}^{T}\boldsymbol{\mu}-\lambda\mathbf{w}^{T}\boldsymbol{\Sigma}\mathbf{w}$ assumes investors have a taste for the quadratic expected utility function on portfolio return. But investors' tastes might be very different.


## From mean-variance portfolio to high-order portfolio

To make up the drawbacks of the mean-variance framework, we need to take high-order moments of the portfolio return into consideration. It admits the following advantages:

- The asymmetry and heavy-tailedness of portfolio return are well captured by its third and fourth central moments, respectively, i.e., skewness and kurtosis.
- Incorporating high-order moments of the portoflio return can approximate a general expected utility function better with its Taylor series expansion truncated to the four most important order terms.

Generally speaking, a higher skewness usually means that the portfolio return admits a more positively skewed shape, while the lower kurtosis usually corresponds to thinner tail. We can extend the mean-variance framework by directly incorporating the high-order moments to obtain the mean-variance-skewness-kurtosis (MVSK) framework, where we shall try to strike a balance between maximizing the mean and skewness (odd moments) while minimizing the variance and kurtosis (even moments).


## High-order portfolios

The package **highOrderPortfolios** implements the algorithms designed in the paper [@zhou2020highOrderPortfolios]. The two main problem formulations covered in that paper are mean-variance-skewness-kurtosis (MVSK) and mean-variance-skewness-kurtosis tilting (MVSKT) problems.

### MVSK portfolio
The simplest formulation for the high-order portfolio is directly extending the Markowitz portfolio objective by considering also the skewness and kurtosis:
$$\begin{aligned}\underset{\mathbf{w}}{\mathsf{minimize}}\,\, & \,\,\,\,\,f\left(\mathbf{w}\right)=-\lambda_{1}\phi_{1}\left(\mathbf{w}\right)+\lambda_{2}\phi_{2}\left(\mathbf{w}\right) -\lambda_{3}\phi_{3}\left(\mathbf{w}\right)+\lambda_{4}\phi_{4}\left(\mathbf{w}\right)\\
\mathsf{subject\,\,to} & \quad\mathbf{w}\in\mathcal{W},
\end{aligned}
$$
where $\lambda_{1},\lambda_{2},\lambda_{3},\lambda_{4}\ge0$ are the parameters for combining four moments of the portfolio return and 

- $\phi_{1}\left(\mathbf{w}\right) = \mathbf{w}^{T}\boldsymbol{\mu}$ is the first moment of portfolio return, 
- $\phi_{2}\left(\mathbf{w}\right) = \mathbf{w}^{T}\boldsymbol{\Sigma}\mathbf{w}$ is the second central moment of portfolio return,
- $\phi_{3}\left(\mathbf{w}\right) = \mathbf{w}^{T}\boldsymbol{\Phi}\left(\mathbf{w}\otimes\mathbf{w}\right)$ is the third central moment of portfolio return, where $\boldsymbol{\Phi} \in \mathbb{R}^{N \times N^2}$ is the co-skewness matrix,
- $\phi_{4}\left(\mathbf{w}\right) = \mathbf{w}^{T}\boldsymbol{\Psi}\left(\mathbf{w}\otimes\mathbf{w}\otimes\mathbf{w}\right)$ is the third central moment of portfolio return, where $\boldsymbol{\Psi} \in \mathbb{R}^{N \times N^3}$ is the co-kurtosis matrix.


### MVSK tilting portfolio
Another formulation for the high-order portfolio is putting the high-order consideration into constraints. We study the algorithm for solving, for example, the MVSK tilting portfolio [@boudt2020mvskt]:
$$\begin{aligned}\underset{\mathbf{w},\delta}{\mathsf{maximize}}\,\, & \quad\delta\\
\mathsf{subject\,\,to} & \quad\phi_{1}\left(\mathbf{w}\right)\ge\phi_{1}\left(\mathbf{w}_{0}\right)+d_{1}\delta,\\
 & \quad\phi_{2}\left(\mathbf{w}\right)\le\phi_{2}\left(\mathbf{w}_{0}\right)-d_{2}\delta,\\
 & \quad\phi_{3}\left(\mathbf{w}\right)\ge\phi_{3}\left(\mathbf{w}_{0}\right)+d_{3}\delta,\\
 & \quad\phi_{4}\left(\mathbf{w}\right)\le\phi_{4}\left(\mathbf{w}_{0}\right)-d_{4}\delta,\\
 & \quad\left(\mathbf{w}-\mathbf{w}_{0}\right)^{T}\boldsymbol{\Sigma}\left(\mathbf{w}-\mathbf{w}_{0}\right)\le\kappa^{2},\\
 & \quad\mathbf{w}\in\mathcal{W},\delta\ge0,
\end{aligned}
$$
where $\mathbf{d}=\left[d_{1},d_{2},d_{3},d_{4}\right]\ge\mathbf{0}$ is the tilting direction, $\kappa^{2}$ determines the maximum tracking error volatility of $\mathbf{w}$ with respect to the reference portfolio $\mathbf{w}_{0}$.

# Using the Package **highOrderPortfolios**

## Estimate the high-order component matrix

Unlike the mean and covariance matrix, the co-skewness and co-kurtosis matrix are rarely estimated. The base R does not provide an embedded function for estimating them. Therefore, we include the function `estimate_moments()` in this package to help estimate the sample co-skewness and co-kurtosis matrix. A simple code example is presented as follows:
```{r, echo=TRUE}
library(highOrderPortfolios)

# load the embedded dataset
data(X50)

# estimate moments
X_moments <- estimate_moments(X50)
names(X_moments)
```


## MVSK portfolio

A simple code example for solving the [MVSK portfolio](#mvsk-portfolio) is presented as follows:

```{r, echo=TRUE}
# decide moment weights
xi <- 10
lmd <- c(1, xi/2, xi*(xi+1)/6, xi*(xi+1)*(xi+2)/24)

# portfolio optimization
sol <- design_MVSK_portfolio(lmd, X_moments)
```

Check the optimized MVSK portfolio:

```{r, echo=TRUE}
data.frame(idx = 1:ncol(X50), weights = sol$w) %>%
  ggplot(aes(x = idx, y = weights)) +
  geom_bar(stat = "identity") +
  labs(x = "Asset indexes", y = "Portfolio weights")
```


## MVSK tilting portfolio
A simple code example for solving the [MVSK tilting portfolio](#mvsk-tilting-portfolio) is presented as follows:

```{r, echo=TRUE}
# estimate moments
X_moments <- estimate_moments(X50, adjust_magnitude = TRUE)

# decide problem setting
w0 <- rep(1/50, 50)
w0_moments <- eval_portfolio_moments(w0, X_moments)
d <- abs(w0_moments) 
kappa <- 0.3 * sqrt(w0 %*% X_moments$Sgm %*% w0)

# portfolio optimization
sol <- design_MVSKtilting_portfolio(d, X_moments, w_init = w0, w0 = w0, 
                                    w0_moments = w0_moments, kappa = kappa)
```

Check the optimized MVSK tilting portfolio:

```{r, echo=TRUE}
sol$improve  # the improvement of moments of the designed portfolio

data.frame(idx = 1:ncol(X50), weights = sol$w) %>%
  ggplot(aes(x = idx, y = weights)) +
  geom_bar(stat = "identity") +
  labs(x = "Asset indexes", y = "Portfolio eights")
```


# References {-}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent
